Gusto: a toolkit for compatible finite element dynamical cores

### 2025-08-04 source/paper.tex
more work on intro

The compatible finite element spaces relevant for geophysical fluid
dynamics are constructed from a discrete differential complex called
the de Rham complex. This is a sequence of spaces with differential
operators between them that map onto the kernel of the next operator
in the sequence. This is often summarised in a commutation diagram as
in figures \ref{fig:2D de Rham} and \ref{fig:3D de Rham}. In these
diagrams the infinite dimensional spaces are along the top row and the
$\pi_i$ denote projections into the corresponding finite dimensional
subspace used in the compatible finite element discretisation. The
differential operators are denoted by $\diff^k$ and these commute with
the projections, meaning that $\diff^{k+1}\pi_k(\cdot) =
\pi_{k+1}\diff^{k+1}(\cdot)$. This guarantees that, in 3D, $\nabla$
maps onto the kernel of $\nabla\times$, ensuring
$\nabla\times\nabla(\cdot)=0$, and $\nabla\times$ maps onto the kernel
of $\nabla\cdot$, ensuring that $\nabla\cdot\nabla\times(\cdot)=0$;
i.e. the spaces are compatible. In 2D we have that $\nabla^\perp$ maps
onto the kernel of $\nabla\cdot$, ensuring that
$\nabla\cdot\nabla^\perp(\cdot)=0$. It is possible to write this de
Rham complex such that instead $\nabla$ maps onto the kernel of
$\nabla^\perp$, ensuring that $\nabla^\perp\cdot\nabla(\cdot)=0$, but
the former is more useful in constructing discretisations for
geophysical fluid dynamics since it enables schemes with local mass
conservation \citep{}.

\begin{figure}\label{fig:2D de Rham}
  \centering{\includegraphics{figures/de_rham_fig_2d.pdf}}
\end{figure}

\begin{figure}\label{fig:3D de Rham}
  \centering{\includegraphics{figures/de_rham_fig_3d.pdf}}
\end{figure}

Due to these promising properties, the next Met Office dynamical core
will use the lowest order compatible finite element discretisation on
the cubed sphere \citep{}. However, there are many important avenues
for research and unanswered questions relating to, for example,
timestepping methods, linear solver techniques and how to couple
unresolved physics processes to the resolved dynamics. This type of
research is difficult in operational models, which are generally
designed to execute specific configurations very efficiently. The
purpose of Gusto is to facilitate this research by providing an open
source, flexible, extensible framework

@@

In this article we will describe the current capabilities of Gusto and
the design features that enable our flexible implementation whereby
different spatial and temporal discretisations can be applied to a
range of different equation sets relevant for weather and climate
modelling. In section \ref{sec: governing} we will outline three
different equations commonly used for modelling geophysical fluids in
order to demonstrate the similarities between the equations and justify
the use of a unified codebase. The advantages of this unified approach
become even clearer in section \ref{sec: FEM} where we present the
compatible finite element discretisation of the different types of
terms (e.g. pressure gradient, transport, diffusion) seen in section
\ref{sec: governing}. In section \ref{sec: design} we describe the
code structure of Gusto followed by a detailed example (section
\ref{sec: FML}) of the Form Manipulation Language (FML) that
facilitates symbolic manipulation of the finite element forms within
our timestepping code. Section \ref{sec: results} contains a selection
of numerical results that showcases the range of discretisations
available within Gusto. **mention here new functionality facilitated
by the flexibility of Gusto? We summarise our current status and
future directions in section \ref{sec: summary}.

@@

\subsection{Compressible Boussinesq Equations}
The compressible Boussinesq equations are commonly used in ocean
modelling. They are useful from a numerical discretisation perspective
because they incorporate compressibility effects without the
complexity of discretising the nonlinear pressure gradient terms that
we shall see in the compressible Euler equations. We write these
equations in terms of the prognostic variables $\MM{u}$, $b$ and $p$
where $\MM{u}$ is the velocity field, $b$ is the fluid buoyancy and
$p$ is pressure. In terms of these variables, the equations are:

\begin{align}
  \pp{\MM{u}}{t} + 
  (\MM{u}\cdot\nabla)\MM{u} +
  2\MM{\Omega}\times \MM{u} + \nabla p + b\hat{\MM{k}} & = 0, \\
  \pp{p}{t} + c^2\nabla\cdot\MM{u} & = 0, \\
  \pp{b}{t} + (\MM{u}\cdot\nabla) b & = 0,
\end{align}
where $\MM{\Omega}$ is the rotation vector of the domain,
$\hat{\MM{k}}$ is the unit vector pointing upwards and $c$ is the
acoustic wave speed. The equations have been written in
three-dimensional form but they can also be solved in a
two-dimensional vertical slice domain described by $(x, z)$
coordinates. In this case the velocity $\MM{u}$ can either be
two-dimensional as well (if there is no Coriolis force), or fully
three-dimensional but with each cartensian component only varying in
the $x-z$ plane. In the latter case this is implemented using a
three-dimensional computational domain that is only one element wide
in the $y$-direction and with periodic boundary conditions also in
that direction.

@@

\begin{align}
  \pp{\MM{u}}{t} + 
  (\MM{u}\cdot\nabla)\MM{u} +
  2\MM{\Omega}\times \MM{u} + c_p\theta\nabla \Pi + g\hat{\MM{k}} & = 0, \\
  \pp{\rho}{t} + \nabla\cdot(\MM{u}\rho) & = 0, \\
  \pp{\theta}{t} + (\MM{u}\cdot\nabla)\theta & = 0, \\
  \Pi^{(1-\kappa)/\kappa} & = \frac{R}{p_0}\rho\theta, & 
\end{align}
where $\Pi$ is the Exner pressure, $p_0$ is a reference pressure, $R$
is the gas constant, $c_p$ is the specific heat at constant pressure
and $\kappa=R/c_p$. The equations have been written in
three-dimensional form but they can also be solved in a
two-dimensional vertical slice domain described by $(x, z)$
coordinates. In this case the velocity $\MM{u}$ can either be two
dimensional as well (if there is no Coriolis force), or fully three
dimensional but with each cartensian component only varying in the
$x-z$ plane. In the latter case this is implemented using a three
dimensional computational domain that is only one element wide in the
$y$-direction and with periodic boundary conditions also in that
direction. \JScomment{I know this repeats what is said above but I
  like that each section is self-contained - I hope that is ok but
  also happy to change it!}

>>>>

The compatible finite element spaces relevant for geophysical fluid
dynamics are constructed from a discrete differential complex called
the de Rham complex. This is a sequence of spaces with differential
operators between them that map onto the kernel of the next operator
in the sequence. This is often summarised in a commutation diagram as
in figures \ref{fig:2DdeRham} and \ref{fig:3DdeRham}. In these
diagrams the infinite dimensional spaces are along the top row and the
$\pi_i$ denote projections into the corresponding finite dimensional
subspace used in the compatible finite element discretisation. The
differential operators are denoted by $\diff^k$ and these commute with
the projections, meaning that $\diff^{k+1}\pi_k(\cdot) =
\pi_{k+1}\diff^{k+1}(\cdot)$. This guarantees that, in 3D, $\nabla$
maps onto the kernel of $\nabla\times$, ensuring
$\nabla\times\nabla(\cdot)=0$, and $\nabla\times$ maps onto the kernel
of $\nabla\cdot$, ensuring that $\nabla\cdot\nabla\times(\cdot)=0$;
i.e. the spaces are compatible. In 2D we have that $\nabla^\perp$
(the 2D curl operator) maps onto the kernel of $\nabla\cdot$,
ensuring that $\nabla\cdot\nabla^\perp(\cdot)=0$. It is possible to
write this de Rham complex such that instead $\nabla$ maps onto the
kernel of $\nabla^\perp$, ensuring that
$\nabla^\perp\cdot\nabla(\cdot)=0$, but the former is more useful in
constructing discretisations for geophysical fluid dynamics since it
enables schemes with local mass conservation \citep{}.

\begin{figure}
  \includegraphics{figures/de_rham_fig_2d.pdf}
  \caption{De Rham complex for two dimensional spaces. The
    interelement continuity of the functions decreases from left to
    right, with functions in $\mathbb{V}^{H^1}$ being fully
    continuous, those in $\mathbb{V}^{HDiv}$ having continuous normal
    components and those in $\mathbb{V}^{L^2}$ being fully
  discontinuous. The polynomial order also decreases from left to
  right.}
  \label{fig:2DdeRham}
\end{figure}

\begin{figure}
  \includegraphics{figures/de_rham_fig_3d.pdf}
  \caption{De Rham complex for three dimensional spaces. The
    interelement continuity of the functions decreases from left to
    right, with functions in $\mathbb{V}^{H^1}$ being fully
    continuous, those in $\mathbb{V}^{HCurl}$ having continuous
    tangential components, those in $\mathbb{V}^{HDiv}$ having
    continuous normal components and those in $\mathbb{V}^{L^2}$ being
    fully discontinuous. The polynomial order also decreases from left
    to right.}
  \label{fig:3DdeRham}
\end{figure}

Due to these promising properties, the next Met Office dynamical core
will use the lowest order compatible finite element discretisation on
the cubed sphere \citep{}. However, there are many important avenues
for research and unanswered questions relating to, for example, the
choice of grid and $\mathbb{V}^{HDiv}_h$; the design of transport
schemes; stable and efficient timestepping algorithms; linear solver
techniques and how to couple unresolved physics processes to the
resolved dynamics. This type of research is difficult in operational
models, which are generally designed to execute specific
configurations very efficiently. The purpose of Gusto is to facilitate
this research by providing an open source, flexible, extensible
framework that provides the key components required when designing
algorithms for a dynamical core. These include: finite element
function spaces defined on the sphere (where the Piola transform is
required to... **); meshes that have a columnar structure (because**),
stable transport and diffusion schemes for variables with different
interelement continuity,

@@

In this article we will describe the current capabilities of Gusto and
the design features that enable our flexible implementation whereby
different spatial and temporal discretisations can be applied to a
range of different equation sets relevant for weather and climate
modelling. In section \ref{sec: governing} we will outline three
different equation sets commonly used for modelling geophysical fluids
in order to demonstrate the similarities between the equation sets and
justify the use of a unified codebase. The advantages of this unified
approach become even clearer in section \ref{sec: FEM} where we
present the compatible finite element discretisation of the different
types of terms (e.g. pressure gradient, transport, diffusion) seen in
section \ref{sec: governing}. In section \ref{sec: design} we describe
the code structure of Gusto followed by a detailed example (section
\ref{sec: FML}) of the Form Manipulation Language (FML) that
facilitates symbolic manipulation of the finite element forms within
our timestepping code. Section \ref{sec: results} contains a selection
of numerical results that showcases the range of discretisations
available within Gusto. We summarise our current status and future
directions in section \ref{sec: summary}.

@@

\subsection{Compressible Boussinesq Equations}
The compressible Boussinesq equations are commonly used in ocean
modelling. They are useful from a numerical discretisation perspective
because they incorporate compressibility effects without the
complexity of discretising the nonlinear pressure gradient terms that
we shall see in the compressible Euler equations. We write these
equations in terms of the prognostic variables $\MM{u}$, $b$ and $p$
where $\MM{u}$ is the velocity field, $p$ is pressure and $b$ is the
fluid buoyancy. In terms of these variables, the equations are:

\begin{align}
  \pp{\MM{u}}{t} + 
  (\MM{u}\cdot\nabla)\MM{u} +
  2\MM{\Omega}\times \MM{u} + \nabla p + b\hat{\MM{k}} & = 0, \\
  \pp{p}{t} + c^2\nabla\cdot\MM{u} & = 0, \\
  \pp{b}{t} + (\MM{u}\cdot\nabla) b & = 0,
\end{align}
where $\MM{\Omega}$ is the rotation vector of the domain,
$\hat{\MM{k}}$ is the unit vector pointing upwards and $c$ is the
acoustic wave speed. The equations describe the motion of a
three dimensional stratified fluid but can also be solved in a
two dimensional vertical slice domain described by $(x, z)$
coordinates. In this case, the velocity $\MM{u}$ can either be
two dimensional as well (if there is no Coriolis force), or fully
three dimensional but with each Cartesian component only varying in
the $x-z$ plane. In the latter case, this is implemented using a
three dimensional computational domain that is only one element wide
in the $y$-direction and with periodic boundary conditions also in
that direction.

@@

\begin{align}
  \pp{\MM{u}}{t} + 
  (\MM{u}\cdot\nabla)\MM{u} +
  2\MM{\Omega}\times \MM{u} + c_p\theta\nabla \Pi + g\hat{\MM{k}} & = 0, \\
  \pp{\rho}{t} + \nabla\cdot(\MM{u}\rho) & = 0, \\
  \pp{\theta}{t} + (\MM{u}\cdot\nabla)\theta & = 0, \\
  \Pi^{(1-\kappa)/\kappa} & = \frac{R}{p_0}\rho\theta, & 
\end{align}
where $\Pi$ is the Exner pressure, $p_0$ is a reference pressure, $R$
is the gas constant, $c_p$ is the specific heat at constant pressure
and $\kappa=R/c_p$. Again, these equations describe the motion of a
three dimensional stratified fluid but can also be solved in a
two dimensional vertical slice domain described by $(x, z)$
coordinates. In this case, the velocity $\MM{u}$ can either be two
dimensional as well (if there is no Coriolis force), or fully three
dimensional but with each Cartesian component only varying in the
$x-z$ plane. In the latter case, this is implemented using a three
dimensional computational domain that is only one element wide in the
$y$-direction and with periodic boundary conditions also in that
direction. \JScomment{I know this repeats what is said above but I
  like that each section is self-contained - I hope that is ok but
  also happy to change it!}
